---
title: "Project: MediCal Eligibility Between Ethncities"
format: html
execute: 
  echo: false
  warning: false
  message: false
embed-resources: true
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(purrr)
library(ggplot2)
library(patchwork)
```

## Data Description

The dataset titled "Race/Ethnicity of Eligible Individuals Under Age 21 Enrolled in Medi-Cal" provides information about the racial and ethnic composition of Medi-Cal eligible individuals aged 0 to 20 in California. The data categories individuals into various races and ethic groups, including Hispanic, White, Black, Asian (Chinese, Filipino, Vietnamese, and others), Alaskan Native or American Indian, and Pacific Islanders. The data is organized by reporting period of year and month, the ethnic group categories, and the number of eligible individuals.

Reference: 

[California Health and Human Services Open Data Portal. (n.d.). Race/Ethnicity of Eligible Individuals Under Age 21 Enrolled in Medi-Cal – Dataset.](https://data.chhs.ca.gov/dataset/eligible-individuals-under-age-21-enrolled-in-medi-cal-by-county/resource/c3bd3780-f419-4369-9f65-9b08cf54cb84) 

```{r}
#| label: Orginal Data
MediCal <- read_csv("data/data_raw/7.4-race-ethnicity-statewide_q4_2024_odp.csv")
```

## Data Cleaning

Data cleaning is an essential first step in preparing the dataset for visualizations. The process begins by loading all the used packages and the original dataset. Then, looking at the dataset, I realized that the names of each column are too long, so I first renamed the variables to shorter and simpler names. After renaming variables, I changed my column of the number of eligible individuals from a string to a numeric type to make sure the data is numeric. Next, I looked at the various ethnic groups in the ethnicity column. I changed one of the groups' names into something more commonly used, such as turning Amerasian into Mixed Asian Americans. Finally, I separated my time into months and years, in case I needed to use separate analyses for the year and month. All of these changes to the original data create my base model for future datasets that I will make, like my dataset based on the total number of eligible individuals or based on 2024 data. 

For the changes I made to get a dataset based on the total number of eligible individuals per year for each ethnic group, I used the clean base model that I created previously. First, I make sure the data is grouped by year and ethnic groups. After this simple arrangement, I turn the number of eligible individuals column into a total column by summarizing the sum of all eligible individuals who share the same ethnicity and year. These two changes create my yearly dataset comprising the year, the different ethnic groups, and the total number of eligible individuals per year for each ethnic group.

I also use the same clean base model for my changes to get the dataset based on recent data, aka the 2024 data. First, I arranged the data into just the year of 2024 by filtering the year section to 2024. Then, I arranged the data based on ethnicity and found the total sums of eligible individuals per ethnic group in the year 2024 by adding up all the monthly data into a new column. These changes created my dataset for 2024.

```{r}
#| label: Data Cleaning

# Making a Base Model of the Clean Data
MediCal <- MediCal |>
  rename(Period = `Reporting Period`, 
         Ethnic = `Race Ethnicity` , 
         Eligiblity = `Number of Eligible Individuals`) |> #Renaming all my variables into something more simpler and understandable
  mutate(Eligiblity = as.numeric(Eligiblity)) |> # Making sure all Eligibility column is numeric data
  mutate(Ethnic = fct_recode(Ethnic, 
                             "Mixed Asian-American" = "Amerasian")) |> # Realized that I need to specify what Amerasian is for those who do not know
    separate(Period, into = c("Year", "Month"), sep = "-") # Want to separate the period into months and year

## write_csv(MediCal, "data/data_clean/MediCal.csv")

# Creating a data set that tells the total of medical eligible individual for each ethnicity per year
Year <- MediCal |>
  group_by(Year, Ethnic) |> 
  summarise(Total = sum(Eligiblity)) # Finding the total number of eligible individuals based on the year and ethnicity

## write_csv(Year, "data/data_clean/Yearly_MediCal.csv")

# 2024 Monthly Total Combined and Separated
Month <- MediCal |>
  filter(Year == "2024") |> 
  group_by(Ethnic) |>
  mutate(Total = sum(Eligiblity))

## write_csv(Month, "data/data_clean/2024Monthly_MediCal.csv")
```

```{r}
#| label: read data
# loading in the clean data here

MediCal_clean <- read_csv("data/data_clean/MediCal.csv")
Yearly <- read_csv("data/data_clean/Yearly_MediCal.csv")
M2024 <- read_csv("data/data_clean/2024Monthly_MediCal.csv")
```


## Visualzation 1

Here is a description of the first visualization. I want to describe what the visualization shows to a general audience. I do not want to see any code or technical terms. I want to be able to read this and understand what the visualization shows. I also want to see a caption and alt text for the visualization. The caption should be a short description of the visualization and the alt text should be a longer description of the visualization. The alt text should be descriptive enough that someone who is blind could understand what the visualization shows.


## Visualization 2

Here is a description of the second visualization. I want to describe what the visualization shows to a general audience. I do not want to see any code or technical terms. I want to be able to read this and understand what the visualization shows. I also want to see a caption and alt text for the visualization. The caption should be a short description of the visualization and the alt text should be a longer description of the visualization. The alt text should be descriptive enough that someone who is blind could understand what the visualization shows.

## Visualization 3

Here is a description of the second visualization. I want to describe what the visualization shows to a general audience. I do not want to see any code or technical terms. I want to be able to read this and understand what the visualization shows. I also want to see a caption and alt text for the visualization. The caption should be a short description of the visualization and the alt text should be a longer description of the visualization. The alt text should be descriptive enough that someone who is blind could understand what the visualization shows.
